Step 2. Screen genomes
==============================================

**Author**: Philipp Muench (pmu15@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
	line-height: 120%;
}
</style>

# search for HMM hits

In this steps we use hmmsearch to find matches of the the HMM models in the `hmm/` folder to the genomic collection in the `genomes/` folder. The `-E` value controls the cutoff for the per-sequence ranked hit list. It can be interpreted as the probability of observing a mapping results by chance.  


```{bash, eval=F}
parallel 'hmmsearch -E 0.0001 --cpu 1 --domtblout {.}.domtblout_lysis hmm/lysis.hmm {}' ::: genomes/faa/*.faa
parallel 'hmmsearch -E 0.0001 --cpu 1 --domtblout {.}.domtblout_colicin hmm/colicin.hmm {}' ::: genomes/faa/*.faa
```

these were submitted as a SGE job to the grid using `qsub -N hmmsearch -M pmu15@helmholtz-hzi.de -m ea -l mem_free=2G -pe smp 20 -cwd hmmer.sh` and took a few minutes. The location of `hmmsearch` at the grid is `/net/metagenomics/projects/colicin-lysis/tools/hmmer-3.2.1/bin/bin/hmmsearch`. 

# process and join the hmmsearch output

It would be too easy that hmmer outputs the results just in a tab-seperated format, instead we need a parser to read in these files...

```{r, eval=F}
iterate_over_dom <- function(folder="genomes/hmmmatches/all_colicin", pattern="\\.domtblout_colicin", name="colicin"){
  require(rhmmer)
  all <- NULL
  lysis_files <- list.files(folder, pattern = pattern, full.names=T)
  pb <- txtProgressBar(min = 0, max = length(lysis_files), style = 3)
  for (i in seq_along(lysis_files)){
    dom <- as.data.frame(read_domtblout(lysis_files[i]))
    if (nrow(dom) > 0){
      dom <- cbind(sample=sub(pattern,"",basename(lysis_files[i])), type=name, dom)
    }
    setTxtProgressBar(pb, i)
  all <- rbind(all, dom)
  }
  close(pb)
  return(all)
}
```

We call that function to iterate over the lysis hmm hits and save the table. 

```{r, eval=F}
colicin_table <- iterate_over_dom(folder="genomes/hmmmatches/all_colicin", pattern="\\.domtblout_colicin", name="colicin")
write.table(colicin_table, "genomes/hmm_matches/colicin.csv", sep=";", quote=F, row.names=F)
```

We do the same for the colicin hits

```{r, eval=F}
lysis_table <- iterate_over_dom(folder="genomes/hmmmatches/all_lysis", pattern="\\.domtblout_lysis", name="lysis")
write.table(lysis_table, "genomes/hmm_matches/lysis.csv", sep=";", quote=F, row.names=F)
```

## Sessioninfo

```{r}
sessionInfo()
```

