Step 2. Plot HMM occurences
==============================================

**Author**: Philipp Muench (pmu15@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
    line-height: 120%;
}
</style>


load libraries

```{r}
rm(list = ls())
library(tidyr)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(gplots)
```

In `2_screen_genomes.Rmd` we created two tables holding the information of the HMM occurence per genome, we import these two tables which are formatted in the `domtblout` format that stores the domain hits table ([reference](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf)). We filter the matches based on the _acc_ column that stores the mean posterior probability of aligned residues in the MEA alignment (from 0 to 1, where 1 indicate a complete alignment). And set this to 0.95. 

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep=";", header=T, stringsAsFactors=F)
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep=";", header=T, stringsAsFactors=F)
merged_all <- rbind(lysis, colicin)
acc_threshold <- 0.90
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```


we replace the HMM description with the Pfam/UniRef ID for three gene families to make sure we can match later these to the `table1.csv` and `table3.csv`

```{r}
merged_all[which(merged_all$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
merged_all[which(merged_all$query_name == "Colicin"),]$query_name <- "UniRef50_A0A0W0TIU2"
merged_all[which(merged_all$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
```

get the data in the wide format (pfam vs. genomes) from the long format (each match to the HMM is one row)

```{r}
lysis$dummy <- 1
lysis_agg <- aggregate(dummy ~ sample, data=lysis, FUN=sum)
colicin$dummy <- 1
colicin_agg <- aggregate(dummy ~ sample, data=colicin, FUN=sum)
merged <- merge(colicin_agg, lysis_agg, by="sample")
colnames(merged) <- c("sample", "colicin", "lysin")
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data=merged_all, FUN=sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
wide_data$sample <- rownames(wide_data)
#chart.Correlation(heatmap_data, histogram=TRUE, pch=19)
```

We prepare the data for plotting and creating a overview figure showing how many genomes (in % of all 9499 genomes) have a match. 

```{r, fig.width=12}
df_m <- melt(wide_data, value.names=sample)
df_m$text <- ""
df_m[which(df_m$value == 0),]$text <- "a 0"
df_m[which(df_m$value == 1),]$text <- "b 1"
df_m[which(df_m$value > 1),]$text <- "c 2-5"
df_m[which(df_m$value > 5),]$text <- "d >5"
df_m$dummy <- 1
df <- aggregate(dummy ~ variable + text, data=df_m, FUN=sum)
# load pfam/UniRef annotations
annot <- read.table("table1.csv", header=T, sep=";")
annot_c <- read.table("table3.csv", header=T, sep=";")
annot_both <- rbind(annot, annot_c)
df$description <- NULL
df$description <- annot_both[match(df$variable, annot_both$ID),]$description
df$group <- annot_both[match(df$variable, annot_both$ID),]$group

# instead of showing the raw number on the x axis, we can show the %
df_percent <- df
df_percent$dummy <- df_percent$dummy / 9499

p <- ggplot(df_percent, aes(x=reorder(description, dummy) , y=dummy, fill=text)) 
p <- p + coord_flip() + facet_grid(group~., space = "free", scale="free")
p <- p + geom_bar(stat="identity") + theme_classic() 
p <- p + scale_fill_manual(values = c("#f7f7f7", "#cccccc", "#969696", "#525252"))
p <- p + scale_y_continuous(labels = scales::percent)
p <- p + labs(x="", y="fraction of genomes", fill="number of HMM matches")
p
```

```{r, fig.width=12}
wide_data.m <- melt(wide_data)
wide_data.m$description <- annot_both[match(wide_data.m$variable, annot_both$ID),]$description
wide_data.m$group <- annot_both[match(wide_data.m$variable, annot_both$ID),]$group

p <- ggplot(wide_data.m, aes(value, fill=group)) + geom_histogram(binwidth=1, color="black") + facet_wrap(group~ description, ncol=5)
p <- p + theme_classic() + xlab("occurences") + ylab("frequency (number of genomes)")
p <- p + theme(strip.text.x = element_text(size = 7))
p <- p + scale_fill_manual(values = c("#f7f7f7", "#cccccc", "#969696", "#525252"))
p <- p + scale_x_continuous(breaks=0:10)
p
```

plotting the number of occurence per genome in a heatmap. 

```{r, fig.width=12, fig.height=12}
wide_data$sample <- NULL
colnames(wide_data) <- annot_both[match(colnames(wide_data), annot_both$ID),]$description
color.palette  <- c("#F8F8F8", colorRampPalette(c("yellow", "orange", "red"))(n=599))
heatmap.2(data.matrix(wide_data),
	density.info="none",
	trace="none",
	labRow = FALSE,
	margins = c(22, 5),
	notecol="black",
	col = color.palette)
```

## Sessioninfo

```{r}
sessionInfo()
```

