Step 2. Plot HMM occurences filtered by _Typhimurium_ match 
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
    line-height: 120%;
}
</style>


load libraries

```{r}
rm(list = ls())
library(tidyr)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(stringr)
library(gplots)
```

In `2_screen_genomes.Rmd` we created two tables holding the information of the HMM occurence per genome, we import these two tables which are formatted in the `domtblout` format that stores the domain hits table ([reference](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf)). We filter the matches based on the _acc_ column that stores the mean posterior probability of aligned residues in the MEA alignment (from 0 to 1, where 1 indicate a complete alignment). And set this to 0.95. 

just take the best match on the colicin

```{r}
acc_threshold <- .7
```

### import colicin hits

```{r}
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$query_name != "Colicin"),] # remove colicin V matches

# iterate over every genome and extract best match
colicin_best <- NULL
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome
colicin[which(colicin$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
colicin <- colicin[which(colicin$acc > acc_threshold),]
```


### import lysis hits

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
lysis[which(lysis$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
lysis <- lysis[which(lysis$acc > acc_threshold),]
```

### merge to one data.frame and filter by accuracy

```{r}
merged_all <- rbind(lysis, colicin)
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```

### reshape data to long format

```{r}
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data = merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

add missing genomes that dont have any hits

```{r}
all_genomes <- read.table("ncbi_list_with_header.csv", header = T, sep = ",")
all_genomes <- all_genomes[which(all_genomes$RefSeq.FTP != ""),]
# add missing genomes
# genome identifier should be something like GCF_000009505.1_ASM950v1_protein
all_genomes$genome_id <- paste0(basename(as.character(all_genomes$RefSeq.FTP)), "_protein")
genomes_match <- match(all_genomes$genome_id, rownames(wide_data))
not_in_wide <- all_genomes[which(is.na(genomes_match)),]
no_hits <- data.frame(matrix(vector(), nrow(not_in_wide), ncol(wide_data),
                dimnames = list(c(), colnames(wide_data))),
                stringsAsFactors = F,
                row.names = not_in_wide$genome_id)
no_hits[is.na(no_hits)] <- 0
wide_data <- rbind(wide_data, no_hits)
```

Show only genomes that belong to _Typhimurium_

```{r}
wide_data$taxa <- all_genomes[match(rownames(wide_data), all_genomes$genome_id),]$Organism.Name
wide_data <- wide_data[grep("Typhimurium", wide_data$taxa),]
taxa <- wide_data$taxa
wide_data$taxa <- NULL
```

We prepare the data for plotting and creating a overview figure showing how many genomes (in % of all 9499 genomes) have a match. 

## only _Typhimurium_

```{r, fig.width=12}
df_m <- melt(wide_data, value.names = sample)
df_m$text <- ""
df_m[which(df_m$value == 0),]$text <- "a 0"
df_m[which(df_m$value == 1),]$text <- "b 1"
df_m[which(df_m$value > 1),]$text <- "c 2-5"
df_m[which(df_m$value > 5),]$text <- "d >5"
df_m$dummy <- 1
df <- aggregate(dummy ~ variable + text, data=df_m, FUN=sum)
# load pfam/UniRef annotations
annot <- read.table("table1.csv", header=T, sep=";")
annot_c <- read.table("table3_curated.csv", header=T, sep=";")
annot_both <- rbind(annot, annot_c)
df$description <- NULL
df$description <- annot_both[match(df$variable, annot_both$ID),]$description
df$group <- annot_both[match(df$variable, annot_both$ID),]$group

# instead of showing the raw number on the x axis, we can show the %
df_percent <- df
df_percent$dummy <- df_percent$dummy / nrow(wide_data)

p <- ggplot(df_percent, aes(x=reorder(description, dummy) , y=dummy, fill=text)) 
p <- p + coord_flip() + facet_grid(group ~ ., space = "free", scale="free")
p <- p + geom_bar(stat="identity") + theme_classic() 
p <- p + scale_fill_manual(values = c("#f7f7f7", "#cccccc", "#969696", "#525252"))
p <- p + scale_y_continuous(labels = scales::percent)
p <- p + labs(x = "", y = "fraction of genomes (Typhimurium)", fill = "number of HMM matches")
p
```


```{r}
pdf("figure2.pdf", width = 7, height = 3.5)
print(p)
dev.off()
```


## only _Typhimurium_ stratified

```{r, fig.width=12}
wide_data$taxa <- word(all_genomes[match(rownames(wide_data), all_genomes$genome_id),]$Organism.Name,-1)

taxa_occ <- as.data.frame(table(wide_data$taxa)) # count occurences
taxa_occ <- taxa_occ[order(-taxa_occ$Freq),][1:3,]  # sort 
wide_data <- wide_data[which(!is.na(match(wide_data$taxa, taxa_occ$Var1))),] # limit to the top 3 names

df_m <- melt(wide_data, value.names = c(sample, taxa))
df_m$text <- ""
df_m[which(df_m$value == 0),]$text <- "a 0"
df_m[which(df_m$value == 1),]$text <- "b 1"
df_m[which(df_m$value > 1),]$text <- "c 2-5"
df_m[which(df_m$value > 5),]$text <- "d >5"
df_m$dummy <- 1
df <- aggregate(dummy ~ variable + text + taxa, data = df_m, FUN = sum)
# load pfam/UniRef annotations
annot <- read.table("table1.csv", header = T, sep = ";")
annot_c <- read.table("table3_curated.csv", header = T, sep = ";")
annot_both <- rbind(annot, annot_c)
df$description <- NULL
df$description <- annot_both[match(df$variable, annot_both$ID),]$description
df$group <- annot_both[match(df$variable, annot_both$ID),]$group

# instead of showing the raw number on the x axis, we can show the %
p <- ggplot(df, aes(x = reorder(description, dummy) , y=dummy, fill=text)) 
p <- p + coord_flip() + facet_grid(group ~ taxa, space = "free", scale="free")
p <- p + geom_bar(stat = "identity") + theme_classic() 
p <- p + scale_fill_manual(values = c("#f7f7f7", "#cccccc", "#969696", "#525252"))
p <- p + labs(x = "", y = "fraction of genomes (Typhimurium)", fill = "number of HMM matches")
p
```

```{r}
pdf("figure3.pdf", width = 12, height = 3.5)
print(p)
dev.off()
```

## Sessioninfo

```{r}
sessionInfo()
```

