Step 2. Plot HMM occurences
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
    line-height: 120%;
}
</style>


load libraries

```{r}
rm(list = ls())
library(tidyr)
library(stringi)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(stringr)
library(gplots)
source("theme.R")
```

In `2_screen_genomes.Rmd` we created two tables holding the information of the HMM occurence per genome, we import these two tables which are formatted in the `domtblout` format that stores the domain hits table ([reference](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf)). We filter the matches based on the _acc_ column that stores the mean posterior probability of aligned residues in the MEA alignment (from 0 to 1, where 1 indicate a complete alignment). And set this to 0.95. 

just take the best match on the colicin

```{r}
acc_threshold <- .7
```

### import colicin hits

```{r}
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$query_name != "Colicin"),] # remove colicin V matches

# iterate over every genome and extract best match
colicin_best <- NULL
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome
colicin[which(colicin$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
colicin <- colicin[which(colicin$acc > acc_threshold),]
```


### import lysis hits

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
lysis[which(lysis$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
lysis <- lysis[which(lysis$acc > acc_threshold),]
```

### merge to one data.frame and filter by accuracy

```{r}
merged_all <- rbind(lysis, colicin)
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```

### reshape data to long format

```{r}
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data = merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

add missing genoems that dont have any hits

```{r}
all_genomes <- read.table("ncbi_list_with_header.csv", header = T, sep = ",")
all_genomes <- all_genomes[which(all_genomes$RefSeq.FTP != ""),]
# add missing genomes
# genome identifier should be something like GCF_000009505.1_ASM950v1_protein
all_genomes$genome_id <- paste0(basename(as.character(all_genomes$RefSeq.FTP)), "_protein")
genomes_match <- match(all_genomes$genome_id, rownames(wide_data))
not_in_wide <- all_genomes[which(is.na(genomes_match)),]
no_hits <- data.frame(matrix(vector(), nrow(not_in_wide), ncol(wide_data),
                dimnames = list(c(), colnames(wide_data))),
                stringsAsFactors = F,
                row.names = not_in_wide$genome_id)
no_hits[is.na(no_hits)] <- 0
wide_data <- rbind(wide_data, no_hits)
```

## most occuring 10 taxa, dont group by occurence

```{r}
n <- 10 #number of taxa to keep
wide_data[wide_data > 0] = 1# binarize
taxas <- all_genomes[match(rownames(wide_data), all_genomes$genome_id),]$Organism.Name
wide_data$taxa <-  gsub("\\ str.*","",taxas)

# remove generic names
wide_data <- wide_data[which(wide_data$taxa != "Salmonella enterica"),]
wide_data <- wide_data[which(wide_data$taxa != "Salmonella enterica subsp. enterica"),]

taxa_occ <- as.data.frame(table(wide_data$taxa))
taxa_occ <- taxa_occ[order(-taxa_occ$Freq),][1:n,]  # sort 
wide_data <- wide_data[which(!is.na(match(wide_data$taxa, taxa_occ$Var1))),] # limit to the top n names

df_m <- melt(wide_data, value.names = c(sample, taxa))
df <- aggregate(value ~ variable + taxa, data = df_m, FUN = sum)
# load pfam/UniRef annotations
annot <- read.table("table1.csv", header = T, sep = ";")
annot_c <- read.table("table3_curated.csv", header = T, sep = ";")
annot_both <- rbind(annot, annot_c)
df$description <- NULL
df$description <- annot_both[match(df$variable, annot_both$ID),]$description
df$group <- annot_both[match(df$variable, annot_both$ID),]$group
df$serovar <- stri_extract_last_words(df$taxa)
df$serovar <- as.character(df$serovar)


df$serovar <- replace(df$serovar, df$serovar == "Typhi", "a-Typhi")
df$serovar <- replace(df$serovar, df$serovar == "Typhimurium", "b-Typhimurium")
df$serovar <- replace(df$serovar, df$serovar == "Enteritidis", "c-Enteritidis")
df$serovar <- replace(df$serovar, df$serovar == "Newport", "e-Newport")

df$serovar <- replace(df$serovar, df$serovar == "Heidelberg", "f-Heidelberg")
df$serovar <- replace(df$serovar, df$serovar == "Dublin", "g-Dublin")
df$serovar <- replace(df$serovar, df$serovar == "Derby", "h-Derby")
df$serovar <- replace(df$serovar, df$serovar == "Weltevreden", "i-Weltevreden")
df$serovar <- replace(df$serovar, df$serovar == "Agona", "j-Agona")
df$serovar <- replace(df$serovar, df$serovar == "Infantis", "k-Infantis")

library(stringi)
library(ggthemes)
library(lemon)

df <- df[which(df$value > 0),]

# annotate occurence

df$occ <- taxa_occ[match(df$taxa, taxa_occ$Var1),]$Freq
df$value_perc <- round(df$value / df$occ * 100)


# instead of showing the raw number on the x axis, we can show the %
p <- ggplot(df, aes(x = reorder(description, -value_perc) , y = value_perc, fill= serovar)) 
p <- p  + facet_grid(.~ group, scales = "free_x", space = "free_y")
#p <- p + scale_y_log10() 
p <- p + geom_col(position = position_dodge2(width = 0.9, preserve = "single")) 
p <- p  + theme_pmuench() 
#p <- p + geom_bar(stat = "identity", position = "dodge")
p <- p + labs(x = "", y = "expressed in genomes (%)", fill = "serovar")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p <- p + scale_fill_Publication() 
p <- p + scale_y_continuous(expand = c(0,0))
pdf("figure4_percent2_log.pdf", width = 7, height = 4)
print(p)
dev.off()
```

## Sessioninfo

```{r}
sessionInfo()
```

