Step 5. Hypothesis testing
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
	line-height: 120%;
}
</style>

# hypothesis

1. Salmonella with prophages (lysis) habour more often colicins
2. Salmonella with prophages (lyisis) habour more often group B colicincs

# analysis 

```{r}
rm(list = ls())
library(reshape2)
library(tidyr)
library("dplyr")
```

## import data and preprocessing

Due to high similarity of colicin genes we only use the best hit per genome if there is any hit.

```{r}
acc_threshold <- .7
```

### import colicin hits

```{r}
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$query_name != "Colicin"),] # remove colicin V matches

# iterate over every genome and extract best match
colicin_best <- NULL
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome
colicin[which(colicin$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
colicin <- colicin[which(colicin$acc > acc_threshold),]
```

### import lysis hits

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
lysis[which(lysis$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
lysis <- lysis[which(lysis$acc > acc_threshold),]
```

### merge to one data.frame and filter by accuracy

```{r}
merged_all <- rbind(lysis, colicin)
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```

### reshape data to long format

```{r}
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data = merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

## count occurences of lysis and colicin, not looking on grouping

current wide_data data.frame only holds genomes that have at least one match of a gene family (n=9226 of 9499 total). For creating contigency tables we have to include 273 genomes that have been screened but don't show any match at all. 

```{r}
all_genomes <- read.table("ncbi_list_with_header.csv", header = T, sep = ",")
all_genomes <- all_genomes[which(all_genomes$RefSeq.FTP != ""),]
# add missing genomes
# genome identifier should be something like GCF_000009505.1_ASM950v1_protein
all_genomes$genome_id <- paste0(basename(as.character(all_genomes$RefSeq.FTP)), "_protein")
genomes_match <- match(all_genomes$genome_id, rownames(wide_data))
not_in_wide <- all_genomes[which(is.na(genomes_match)),]
no_hits <- data.frame(matrix(vector(), nrow(not_in_wide), ncol(wide_data),
                dimnames = list(c(), colnames(wide_data))),
                stringsAsFactors = F,
                row.names = not_in_wide$genome_id)
no_hits[is.na(no_hits)] <- 0
wide_data <- rbind(wide_data, no_hits)
```


```{r}
# read in annotation tables
lysis <- read.table("table1.csv", header = T, sep = ";")
colicin <- read.table("table3_curated.csv", header = T, sep = ";")

subset_lysis <- rowSums(select(wide_data, starts_with("P"))) > 0
subset_colicin <- rowSums(select(wide_data, starts_with("UniRef50"))) > 0

print(paste("number of genomes with lysis matches", length(which(subset_lysis))))
print(paste("number of genomes without lysis matches", length(which(!subset_lysis))))

print(paste("number of genomes with colicin matches", length(which(subset_colicin))))
print(paste("number of genomes without colicin matches", length(which(!subset_colicin))))

# locical that holds the information which genomes contain a lysis gene
lyis_subset <- rowSums(select(wide_data, lysis$ID)) > 0
nonlyis_subset <- rowSums(select(wide_data, lysis$ID)) == 0

# subset wide_data for one that have/not have a lysis gene
wide_data_w_lysis <- wide_data[lyis_subset,]
wide_data_wo_lysis <- wide_data[nonlyis_subset,]

# subset for presence of colicin
wide_data_w_lysis_w_colicin <-  wide_data[which(subset_lysis & subset_colicin),]
wide_data_w_lysis_wo_colicin <- wide_data[which(subset_lysis & !subset_colicin),]
wide_data_wo_lysis_w_colicin <- wide_data[which(!subset_lysis & subset_colicin),]
wide_data_wo_lysis_wo_colicin <- wide_data[which(!subset_lysis & !subset_colicin),]

# create matrix for testing
contigency_table <-
matrix(c(nrow(wide_data_w_lysis_w_colicin),
         nrow(wide_data_wo_lysis_w_colicin),
         nrow(wide_data_w_lysis_wo_colicin),
         nrow(wide_data_wo_lysis_wo_colicin)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin", "without colicin")))
contigency_table


lysis_sub <- wide_data[which(!subset_lysis),] 
lysis_sub$taxa <- all_genomes[match(rownames(lysis_sub), all_genomes$genome_id),]$Organism.Name
View(as.data.frame(table(lysis_sub$taxa)))
```

```{r, echo=F}
print(paste0("Fraction of bacteria that have lysis (one or more lysis related Pfams): ",
            round(100*(nrow(wide_data_w_lysis_w_colicin) + 
                         nrow(wide_data_w_lysis_wo_colicin)) 
                  / nrow(wide_data), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis (no lysis related Pfams): ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin) + 
                         nrow(wide_data_wo_lysis_wo_colicin)) 
                  / nrow(wide_data), digits = 2),"%"))
print(paste0("Fraction of bacteria that have lysis AND a colicin gene: ",
            round(100*(nrow(wide_data_w_lysis_w_colicin)) 
                  / (nrow(wide_data_w_lysis_w_colicin) + 
                       nrow(wide_data_w_lysis_wo_colicin)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin)) 
                  / (nrow(wide_data_wo_lysis_w_colicin) + 
                       nrow(wide_data_wo_lysis_wo_colicin)), digits = 2),"%"))
```

```{r}
ftest <- fisher.test(contigency_table, alternative = "greater")
print(ftest)
```


```{r, echo=F}
print(paste0("The presence of a colocin gene is significantly enriched in bacteria that have a lysis gene (", 
             nrow(wide_data_w_lysis_w_colicin), " out of ",  nrow(wide_data_w_lysis_w_colicin) + nrow(wide_data_w_lysis_wo_colicin),
      ") compared to bacteria without a lysis gene (",
      nrow(wide_data_wo_lysis_w_colicin), " have a colicin gene out of ",  nrow(wide_data_wo_lysis_w_colicin) + nrow(wide_data_wo_lysis_wo_colicin),
      ") (one-sided Fisher's exact test p-value = ", ftest$p.value,
      ")"))
```

## stratify by colicin group

```{r}
colicin <- read.table("table3_curated.csv", header = T, sep = ";")
colicin_a <- colicin[which(colicin$group == "A"),]
colicin_b <- colicin[which(colicin$group == "B"),]

# subset again for presence of colicin group A and B
# subset that have lysis

subset_lysis <- rowSums(select(wide_data, starts_with("P"))) > 0

print(paste("number of genomes with lysis matches", length(which(subset_lysis))))
print(paste("number of genomes without lysis matches", length(which(!subset_lysis))))

subset_colicin_a <- rowSums(select(wide_data, UniRef50_P02978, UniRef50_P09883, UniRef50_Q47502, UniRef50_P08083)) > 0
print(paste("number of genomes with colicin a matches", length(which(subset_colicin_a))))
print(paste("number of genomes without colicin a matches", length(which(!subset_colicin_a))))

subset_colicin_b <- rowSums(select(wide_data, UniRef50_P05819, UniRef50_P17998, UniRef50_P06716, UniRef50_P05820)) > 0
print(paste("number of genomes with colicin b matches", length(which(subset_colicin_b))))
print(paste("number of genomes without colicin b matches", length(which(!subset_colicin_b))))

# subset genomes that have lysis
wide_data_w_lysis_w_colicin_a <- wide_data[which(subset_lysis & subset_colicin_a),]
wide_data_w_lysis_w_colicin_b <- wide_data[which(subset_lysis & subset_colicin_b),]
wide_data_w_lysis_wo_colicin_a <- wide_data[which(subset_lysis & !subset_colicin_a),]
wide_data_w_lysis_wo_colicin_b <- wide_data[which(subset_lysis & !subset_colicin_b),]

# subset genomes that don't have lysis
wide_data_wo_lysis_w_colicin_a <- wide_data[which(!subset_lysis & subset_colicin_a),]
wide_data_wo_lysis_w_colicin_b <- wide_data[which(!subset_lysis & subset_colicin_b),]
wide_data_wo_lysis_wo_colicin_a <- wide_data[which(!subset_lysis & !subset_colicin_a),]
wide_data_wo_lysis_wo_colicin_b <- wide_data[which(!subset_lysis & !subset_colicin_b),]

# create matrix for testing
contigency_table_colicin_b <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_b),
         nrow(wide_data_wo_lysis_w_colicin_b),
         nrow(wide_data_w_lysis_wo_colicin_b),
         nrow(wide_data_wo_lysis_wo_colicin_b)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_b", "without colicin_b")))
contigency_table_colicin_b

contigency_table_colicin_a <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_a),
         nrow(wide_data_wo_lysis_w_colicin_a),
         nrow(wide_data_w_lysis_wo_colicin_a),
         nrow(wide_data_wo_lysis_wo_colicin_a)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_a", "without colicin_a")))
contigency_table_colicin_a
```

```{r, echo=F}
print(paste0("Fraction of bacteria that have lysis AND a colicin B gene among bacteria taht have a lysis gene ",
            round(100*(nrow(wide_data_w_lysis_w_colicin_b)) 
                  / (nrow(wide_data_w_lysis_w_colicin_b) + 
                       nrow(wide_data_w_lysis_wo_colicin_b)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin B gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin_b)) 
                  / (nrow(wide_data_wo_lysis_w_colicin_b) + 
                       nrow(wide_data_wo_lysis_wo_colicin_b)), digits = 2),"%"))
print(paste0("Fraction of bacteria that have lysis AND a colicin A gene among bacteria taht have a lysis gene ",
            round(100*(nrow(wide_data_w_lysis_w_colicin_a)) 
                  / (nrow(wide_data_w_lysis_w_colicin_a) + 
                       nrow(wide_data_w_lysis_wo_colicin_a)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin A gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin_a)) 
                  / (nrow(wide_data_wo_lysis_w_colicin_a) + 
                       nrow(wide_data_wo_lysis_wo_colicin_a)), digits = 2),"%"))
```

```{r}
ftest_colicin_b <- fisher.test(contigency_table_colicin_b, alternative = "greater")
print(ftest_colicin_b)

ftest_colicin_a <- fisher.test(contigency_table_colicin_a, alternative = "greater")
print(ftest_colicin_a)
```

```{r}
contigency_table_colicin_ab <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_b),
         nrow(wide_data_wo_lysis_w_colicin_b),
         nrow(wide_data_w_lysis_w_colicin_a),
         nrow(wide_data_wo_lysis_w_colicin_a)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_b", "with colicin_a")))
contigency_table_colicin_ab

ftest_colicin_ab <- fisher.test(contigency_table_colicin_b, alternative = "greater")
print(ftest_colicin_ab)
```




## Sessioninfo

```{r}
sessionInfo()
```


