Step 5. Hypothesis testing
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
	line-height: 120%;
}
</style>

# hypothesis

1. Salmonella with prophages (lysis) habour more often colicins
2. Salmonella with prophages (lyisis) habour more often group B colicincs

# analysis 

## import data and preprocessing

Due to high similarity of colicin genes we only use the best hit per genome if there is any hit.

```{r}
rm(list = ls())
library(reshape2)
library(tidyr)
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$domain_name != "WP_000262102.1"),]
#sample <- "GCF_000962315.1_ASM96231v1_protein"
colicin_best <- NULL
# iterate over every genome
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
 # print(paste("sample: ", sample, " found matches: " , nrow(subset)))
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome

lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
merged_all <- rbind(lysis, colicin)
acc_threshold <- 0.70
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
# rename
merged_all[which(merged_all$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
merged_all[which(merged_all$query_name == "Colicin"),]$query_name <- "UniRef50_A0A0W0TIU2"
merged_all[which(merged_all$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
# reshape
lysis$dummy <- 1
lysis_agg <- aggregate(dummy ~ sample, data = lysis, FUN = sum)
colicin$dummy <- 1
colicin_agg <- aggregate(dummy ~ sample, data = colicin, FUN = sum)
merged <- merge(colicin_agg, lysis_agg, by = "sample")
colnames(merged) <- c("sample", "colicin", "lysin")
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data=merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

## count occurences of lysis and colicin, not looking on grouping

```{r}
library("dplyr")
# get annotation of lysis genes
# drop omni-present colicin V
wide_data <- select(wide_data, -"UniRef50_A0A0W0TIU2")

lysis <- read.table("table1.csv", header = T, sep = ";")
colicin <- read.table("table3_curated.csv", header = T, sep = ";")

# locical that holds the information which genomes contain a lysis gene
lyis_subset <- rowSums(select(wide_data, lysis$ID)) > 0
nonlyis_subset <- rowSums(select(wide_data, lysis$ID)) == 0

# subset wide_data for one that have a lysis gene
wide_data_w_lysis <- wide_data[lyis_subset,]
# subset again for presence of colicin
wide_data_w_lysis_w_colicin <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin$ID)) > 0),]
wide_data_w_lysis_wo_colicin <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin$ID)) == 0),]

# subset wide_data for one that dont have a lysis gene
wide_data_wo_lysis <- wide_data[nonlyis_subset,]
# subset again for presence of colicin
wide_data_wo_lysis_w_colicin <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin$ID)) > 0),]
wide_data_wo_lysis_wo_colicin <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin$ID)) == 0),]

# create matrix for testing
contigency_table <-
matrix(c(nrow(wide_data_w_lysis_w_colicin),
         nrow(wide_data_wo_lysis_w_colicin),
         nrow(wide_data_w_lysis_wo_colicin),
         nrow(wide_data_wo_lysis_wo_colicin)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin", "without colicin")))
contigency_table
```

```{r, echo=F}
print(paste0("Fraction of bacteria that have lysis (one or more lysis related Pfams): ",
            round(100*(nrow(wide_data_w_lysis_w_colicin) + 
                         nrow(wide_data_w_lysis_wo_colicin)) 
                  / nrow(wide_data), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis (no lysis related Pfams): ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin) + 
                         nrow(wide_data_wo_lysis_wo_colicin)) 
                  / nrow(wide_data), digits = 2),"%"))
print(paste0("Fraction of bacteria that have lysis AND a colicin gene: ",
            round(100*(nrow(wide_data_w_lysis_w_colicin)) 
                  / (nrow(wide_data_w_lysis_w_colicin) + 
                       nrow(wide_data_w_lysis_wo_colicin)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin)) 
                  / (nrow(wide_data_wo_lysis_w_colicin) + 
                       nrow(wide_data_wo_lysis_wo_colicin)), digits = 2),"%"))
```

```{r}
ftest <- fisher.test(contigency_table, alternative = "greater")
print(ftest)
```


```{r, echo=F}
print(paste0("The presence of a colocin gene is significantly enriched in bacteria that have a lysis gene (", 
             nrow(wide_data_w_lysis_w_colicin), " out of ",  nrow(wide_data_w_lysis_w_colicin) + nrow(wide_data_w_lysis_wo_colicin),
      ") compared to bacteria without a lysis gene (",
      nrow(wide_data_wo_lysis_w_colicin), " have a colicin gene out of ",  nrow(wide_data_wo_lysis_w_colicin) + nrow(wide_data_wo_lysis_wo_colicin),
      ") (one-sided Fisher's exact test p-value = ", ftest$p.value,
      ")"))
```

## stratify by colicin group

```{r}
colicin <- read.table("table3_curated.csv", header = T, sep = ";")
colicin_a <- colicin[which(colicin$group == "A"),]
colicin_b <- colicin[which(colicin$group == "B"),]

# subset again for presence of colicin group A and B
# subset that have lysis
wide_data_w_lysis_w_colicin_a <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin_a$ID)) > 0),]
wide_data_w_lysis_w_colicin_b <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin_b$ID)) > 0),]

wide_data_w_lysis_wo_colicin_a <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin_a$ID)) == 0),]
wide_data_w_lysis_wo_colicin_b <- wide_data_w_lysis[which(rowSums(select(wide_data_w_lysis, colicin_b$ID)) == 0),]

# subset that miss lysis
wide_data_wo_lysis_w_colicin_a <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin_a$ID)) > 0),]
wide_data_wo_lysis_w_colicin_b <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin_b$ID)) > 0),]

wide_data_wo_lysis_wo_colicin_a <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin_a$ID)) == 0),]
wide_data_wo_lysis_wo_colicin_b <- wide_data_wo_lysis[which(rowSums(select(wide_data_wo_lysis, colicin_b$ID)) == 0),]

# create matrix for testing
contigency_table_colicin_b <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_b),
         nrow(wide_data_wo_lysis_w_colicin_b),
         nrow(wide_data_w_lysis_wo_colicin_b),
         nrow(wide_data_wo_lysis_wo_colicin_b)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_b", "without colicin_b")))
contigency_table_colicin_b


contigency_table_colicin_a <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_a),
         nrow(wide_data_wo_lysis_w_colicin_a),
         nrow(wide_data_w_lysis_wo_colicin_a),
         nrow(wide_data_wo_lysis_wo_colicin_a)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_a", "without colicin_a")))
contigency_table_colicin_a
```

```{r, echo=F}
print(paste0("Fraction of bacteria that have lysis AND a colicin B gene among bacteria taht have a lysis gene ",
            round(100*(nrow(wide_data_w_lysis_w_colicin_b)) 
                  / (nrow(wide_data_w_lysis_w_colicin_b) + 
                       nrow(wide_data_w_lysis_wo_colicin_b)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin B gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin_b)) 
                  / (nrow(wide_data_wo_lysis_w_colicin_b) + 
                       nrow(wide_data_wo_lysis_wo_colicin_b)), digits = 2),"%"))
print(paste0("Fraction of bacteria that have lysis AND a colicin A gene among bacteria taht have a lysis gene ",
            round(100*(nrow(wide_data_w_lysis_w_colicin_a)) 
                  / (nrow(wide_data_w_lysis_w_colicin_a) + 
                       nrow(wide_data_w_lysis_wo_colicin_a)), digits = 2),"%"))
print(paste0("Fraction of bacteria that dont have lysis AND a colicin A gene among bacteria without lysis ",
            round(100*(nrow(wide_data_wo_lysis_w_colicin_a)) 
                  / (nrow(wide_data_wo_lysis_w_colicin_a) + 
                       nrow(wide_data_wo_lysis_wo_colicin_a)), digits = 2),"%"))
```

```{r}
ftest_colicin_b <- fisher.test(contigency_table_colicin_b, alternative = "greater")
print(ftest_colicin_b)
```

## Sessioninfo

```{r}
sessionInfo()
```


