Step 5. Hypothesis testing
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
	line-height: 120%;
}
</style>

load libraries

```{r}
rm(list = ls())
library(tidyr)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(stringr)
library(gplots)
library(reshape2)
library(tidyr)
library("dplyr")
```

In `2_screen_genomes.Rmd` we created two tables holding the information of the HMM occurence per genome, we import these two tables which are formatted in the `domtblout` format that stores the domain hits table ([reference](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf)). We filter the matches based on the _acc_ column that stores the mean posterior probability of aligned residues in the MEA alignment (from 0 to 1, where 1 indicate a complete alignment). And set this to 0.95. 

just take the best match on the colicin

```{r}
acc_threshold <- .7
```

### import colicin hits

```{r}
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$query_name != "Colicin"),] # remove colicin V matches

# iterate over every genome and extract best match
colicin_best <- NULL
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome
colicin[which(colicin$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
colicin <- colicin[which(colicin$acc > acc_threshold),]
```


### import lysis hits

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
lysis[which(lysis$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
lysis <- lysis[which(lysis$acc > acc_threshold),]
```

### merge to one data.frame and filter by accuracy

```{r}
merged_all <- rbind(lysis, colicin)
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```

### reshape data to long format

```{r}
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data = merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

add missing genoems that dont have any hits

```{r}
all_genomes <- read.table("ncbi_list_with_header.csv", header = T, sep = ",")
all_genomes <- all_genomes[which(all_genomes$RefSeq.FTP != ""),]
# add missing genomes
# genome identifier should be something like GCF_000009505.1_ASM950v1_protein
all_genomes$genome_id <- paste0(basename(as.character(all_genomes$RefSeq.FTP)), "_protein")
genomes_match <- match(all_genomes$genome_id, rownames(wide_data))
not_in_wide <- all_genomes[which(is.na(genomes_match)),]
no_hits <- data.frame(matrix(vector(), nrow(not_in_wide), ncol(wide_data),
                dimnames = list(c(), colnames(wide_data))),
                stringsAsFactors = F,
                row.names = not_in_wide$genome_id)
no_hits[is.na(no_hits)] <- 0
wide_data <- rbind(wide_data, no_hits)
```

filter to _Typhimurium_

```{r}
wide_data$taxa <- all_genomes[match(rownames(wide_data), all_genomes$genome_id),]$Organism.Name
wide_data <- wide_data[grep("Typhimurium", wide_data$taxa),]
taxa <- wide_data$taxa
wide_data$taxa <- NULL
```

```{r}
colicin <- read.table("table3_curated.csv", header = T, sep = ";")
colicin_a <- colicin[which(colicin$group == "A"),]
colicin_b <- colicin[which(colicin$group == "B"),]

# subset again for presence of colicin group A and B
# subset that have lysis

subset_lysis <- rowSums(select(wide_data, starts_with("P"))) > 0

print(paste("number of genomes with lysis matches", length(which(subset_lysis))))
print(paste("number of genomes without lysis matches", length(which(!subset_lysis))))

subset_colicin_a <- rowSums(select(wide_data, UniRef50_P02978,
                                   UniRef50_P09883,
                                   UniRef50_Q47502,
                                   UniRef50_P08083)) > 0
print(paste("number of genomes with colicin a matches", length(which(subset_colicin_a))))
print(paste("number of genomes without colicin a matches", length(which(!subset_colicin_a))))

subset_colicin_b <- rowSums(select(wide_data, 
                                   UniRef50_P05819, 
                                   UniRef50_P17998, 
                                   UniRef50_P06716, 
                                   UniRef50_P05820)) > 0
print(paste("number of genomes with colicin b matches", length(which(subset_colicin_b))))
print(paste("number of genomes without colicin b matches", length(which(!subset_colicin_b))))

# subset genomes that have lysis
wide_data_w_lysis_w_colicin_a <- wide_data[which(subset_lysis & subset_colicin_a),]
wide_data_w_lysis_w_colicin_b <- wide_data[which(subset_lysis & subset_colicin_b),]
wide_data_w_lysis_wo_colicin_a <- wide_data[which(subset_lysis & !subset_colicin_a),]
wide_data_w_lysis_wo_colicin_b <- wide_data[which(subset_lysis & !subset_colicin_b),]

# subset genomes that don't have lysis
wide_data_wo_lysis_w_colicin_a <- wide_data[which(!subset_lysis & subset_colicin_a),]
wide_data_wo_lysis_w_colicin_b <- wide_data[which(!subset_lysis & subset_colicin_b),]
wide_data_wo_lysis_wo_colicin_a <- wide_data[which(!subset_lysis & !subset_colicin_a),]
wide_data_wo_lysis_wo_colicin_b <- wide_data[which(!subset_lysis & !subset_colicin_b),]

# create matrix for testing

contigency_table_colicin_ab <-
matrix(c(nrow(wide_data_w_lysis_w_colicin_b),
         nrow(wide_data_wo_lysis_w_colicin_b),
         nrow(wide_data_w_lysis_w_colicin_a),
         nrow(wide_data_wo_lysis_w_colicin_a)),
       nrow = 2,
       dimnames = list(colicin = c("with lysis", "without lysis"),
                       lysis = c("with colicin_b", "with colicin_a")))
contigency_table_colicin_ab
```


```{r}
ftest <- fisher.test(contigency_table_colicin_ab)
print(ftest)
```

## Sessioninfo

```{r}
sessionInfo()
```


