Step 3b. Plot HMM occurences as heatmap
==============================================

**Author**: Philipp Muench (philipp.muench@helmholtz-hzi.de)

**Date**: `r Sys.Date()`

----------------------------------------

<style type="text/css">
h1 {
    line-height: 120%;
}
</style>


load libraries

```{r}
rm(list = ls())
library(tidyr)
library(stringi)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(stringr)
library(gplots)
library(wesanderson)
```


In `2_screen_genomes.Rmd` we created two tables holding the information of the HMM occurence per genome, we import these two tables which are formatted in the `domtblout` format that stores the domain hits table ([reference](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf)). We filter the matches based on the _acc_ column that stores the mean posterior probability of aligned residues in the MEA alignment (from 0 to 1, where 1 indicate a complete alignment). And set this to 0.95. 

just take the best match on the colicin

```{r}
acc_threshold <- .7
```

### import colicin hits

```{r}
colicin <- read.table("genomes/hmm_matches/colicin.csv", sep = ";", header = T, stringsAsFactors = F)
colicin <- colicin[which(colicin$query_name != "Colicin"),] # remove colicin V matches

# iterate over every genome and extract best match
colicin_best <- NULL
for (sample in unique(colicin$sample)) {
  subset <- colicin[which(colicin$sample == sample),]
  hit <- which.min(subset$sequence_evalue)
  colicin_best <- rbind(colicin_best, subset[hit,])
}
colicin <- colicin_best # now this table is limited to one match per genome
colicin[which(colicin$query_name == "Colicin-E1"),]$query_name <- "UniRef50_P02978"
colicin <- colicin[which(colicin$acc > acc_threshold),]
```


### import lysis hits

```{r}
lysis <- read.table("genomes/hmm_matches/lysis.csv", sep = ";", header = T, stringsAsFactors = F)
lysis[which(lysis$query_name == "Phage_lysozyme"),]$query_name <- "PF00959"
lysis <- lysis[which(lysis$acc > acc_threshold),]
```

### merge to one data.frame and filter by accuracy

```{r}
merged_all <- rbind(lysis, colicin)
merged_all <- merged_all[which(merged_all$acc > acc_threshold),]
```

### reshape data to long format

```{r}
merged_all$dummy <- 1
wide_data <- aggregate(dummy ~ sample + query_name, data = merged_all, FUN = sum)
wide_data <- spread(wide_data, query_name, dummy)
wide_data[is.na(wide_data)] <- 0
rownames(wide_data) <- wide_data$sample
wide_data$sample <- NULL
```

add missing genomes that dont have any hits

```{r}
all_genomes <- read.table("ncbi_list_with_header.csv", header = T, sep = ",")
all_genomes <- all_genomes[which(all_genomes$RefSeq.FTP != ""),]
# add missing genomes
# genome identifier should be something like GCF_000009505.1_ASM950v1_protein
all_genomes$genome_id <- paste0(basename(as.character(all_genomes$RefSeq.FTP)), "_protein")
genomes_match <- match(all_genomes$genome_id, rownames(wide_data))
not_in_wide <- all_genomes[which(is.na(genomes_match)),]
no_hits <- data.frame(matrix(vector(), nrow(not_in_wide), ncol(wide_data),
                dimnames = list(c(), colnames(wide_data))),
                stringsAsFactors = F,
                row.names = not_in_wide$genome_id)
no_hits[is.na(no_hits)] <- 0
wide_data <- rbind(wide_data, no_hits)
```


```{r}
# load pfam/UniRef annotations
annot <- read.table("table1.csv", header = T, sep = ";")
annot_c <- read.table("table3_curated.csv", header = T, sep = ";")
annot_both <- rbind(annot, annot_c)
```

use `complex.heatmap` to plot

```{r}
library(ComplexHeatmap)
library(circlize)
wide_data$sample <- NULL
annotation_data <- data.frame(type = annot_both[match(colnames(wide_data), annot_both$ID),]$group)
colnames(wide_data) <- annot_both[match(colnames(wide_data), annot_both$ID),]$description
```

get taxa information to the right shape

```{r}
taxa <- all_genomes[match(rownames(wide_data), all_genomes$genome_id),]$Organism.Name
taxa <-  gsub("\\ str.*","",taxa)
serovar <- stri_extract_last_words(taxa)
serovar_list <- c("Typhi",
                  "Typhimurium",
                  "Enteritidis",
                  "Newport",
                  "Heidelberg",
                  "Dublin",
                  "Derby",
                  "Weltevreden",
                  "Agona",
                  "Infantis")

serovar <- stri_extract_last_words(taxa)
serovar[which(!serovar %in% serovar_list)] <- "other"

```


```{r, fig.width=12, fig.height=12}
ha = HeatmapAnnotation(type = as.character(annotation_data$type),
    col = list(type = c("lysin" = "#b2df8a", "A" = "#a6cee3", "B" = "#1f78b4", "none" = "grey")))
ha2 = HeatmapAnnotation(barplot = anno_barplot(colSums(wide_data), bar_width = 1, gp = gpar(col = NA, fill = "grey"), 
    border = FALSE, axis = TRUE))

taxa_type <- rep("other", nrow(wide_data))
taxa_data = word(taxa,-1)
taxa_type[which(taxa_data == "DT104")] <- "DT104"
taxa_type[which(taxa_data == "5-")] <- "5-"

pal <- wes_palette("Zissou1", 100, type = "continuous")


ht <- Heatmap(name = "heat", top_annotation = ha2,
        col = colorRamp2(c(0, 1, 5, 9), c("grey95", "grey50", "grey20", "black")),
        top_annotation_height = unit(1.5, "cm"),
        bottom_annotation_height = unit(.5, "cm"),
        bottom_annotation = ha,
        wide_data,
        column_order = order(as.character(annotation_data$type)),
        cluster_columns = FALSE,
        cluster_rows = TRUE,
        show_row_dend = FALSE,
        row_names_side = "left",
        show_row_names = FALSE) + Heatmap(serovar, name = "subtype", col = c("Typhi" = "#386cb0", 
                                                                            "Agona" = "#ffff33", 
                                                                            "Enteritidis" = "#7fc97f",
                                                                            "Infantis" = "black",
                                                                            "Newport" = "#ef3b2c",
                                                                            "Typhimurium" = "#fdb462",
                                                                            "Derby" = "#fb9a99",
                                                                            "Heidelberg" = "#662506",
                                                                            "Weltevreden" = "#984ea3",
                                                                            "Dublin" = "#a6cee3",
                                                                            "other" = "grey"
                                                                            ), 
    width = unit(0.5, "cm"))


pdf(file = "ht2.pdf", width = 4, height = 6)
print(ht)
dev.off()
```